### 2023-05-02 tex/linear.tex
[ edit ] new draft of chapters 5 and 6
[ writing ] mostly C3 related work

\chapter{Linearity}\label{sec:linearity}

@@

\section{DILL}
\input{dill.tex}

\section{Mechanisation survey}

>>>>

\chapter{Linearity and modality}\label{sec:linearity}

\begin{itemize}
  \item Motivation via prior work:
    \begin{itemize}
      \item Rust
      \item Session types
      \item ATS
      \item Complexity \citep{Hofmann03}
      \item Modal\ldots
    \end{itemize}
  \item Can \citet{AACMM21} capture linearity/modality?
    It's unclear based on standard presentations.
    Let's find one that looks good (DILL/PD), and see whether it works.
  \item In-depth introduction of DILL and PD
  \item It's close to working, but doesn't.
  \item Mention \citet{BBdePH93}, problems with substitution.
\end{itemize}

The techniques of \cref{sec:simple} were developed to handle idealised core
calculi, like the simply typed $\lambda$-calculus and its many variants.
Such core calculi are fairly close to some real intermediate languages,
\todo{cite} though many intermediate languages also have polymorphism in their
type systems, which I have not discussed, but has been considered by much prior
work\todo{cite}.

%Linear logic was introduced by \citet{girard87linear} as a tool of proof theory.
%Its distinguishing feature is that hypotheses, by default, must be used exactly
%once.
%I devote a chapter to introducing linear logic because it is better understood
%than any individual or general semiring-annotated calculi, and is thus an
%important point of comparison.
%Linear logic will be an instance of the semiring-annotated systems we study
%later.

@@

\section{DILL}\label{sec:dill}
\input{dill.tex}

\section{Mechanisation survey}
\input{linmech.tex}

### 2023-05-02 tex/lrkits.tex
[ edit ] another draft of chapters 5 and 6
[ writing ] mostly C3 related work

\begin{example}
  Take $\Ann = \plr{\mathbb N, =, 0, +, 1, \times}$, with the equality order
  chosen to avoid any concerns around subsumption of annotations.
  Then, there is an intuitionistic recursive environment as follows, where
  $y\,z$ is the application of $y$ to $z$.
  \[
    \alr{\alr{z},\alr{y\,z}} :
    \plr{x : A, y : B \to C, z : B} \env\vdash_R \plr{B, C}.
  \]
  There is also a usage-aware recursive environment
  \[
    \alr{\alr{z},\alr{y\,z}} :
    \plr{\gr0x : A, \gr2y : B \multimap C, \gr3z : B} \env\vdash_R
    \plr{\gr1B, \gr2C}.
  \]
  The latter relies on the observations that
  $\begin{pmatrix} \gr0 & \gr2 & \gr3 \end{pmatrix} =
  \begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix}
  + \begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix}$ and, on the right, that
  $\begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix} =
  \gr2\begin{pmatrix} \gr0 & \gr1 & \gr1 \end{pmatrix}$.
  Then, we have $\gr0x : A, \gr0y : B \multimap C, \gr1z : B \vdash z : B$ and
  $\gr0x : A, \gr1y : B \multimap C, \gr1z : B \vdash y\,z : C$.
\end{example}

From the example, we can see that the important usage vectors are the initial
one $\begin{pmatrix} \gr0 & \gr2 & \gr3 \end{pmatrix}$ and the usage vectors
at which terms are derived: $\begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix}$
and $\begin{pmatrix} \gr0 & \gr1 & \gr1 \end{pmatrix}$.
I will call the latter the \emph{leaf vectors}.
The intermediate vector $\begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix}$ can
be worked out from the leaf vector
$\begin{pmatrix} \gr0 & \gr1 & \gr1 \end{pmatrix}$ and the scaling factor
$\gr2$ found in the codomain context $\gr1B, \gr2C$.
Even when the ordering on annotations is given by a non-equivalence relation
$\leq$, there is a canonical least choice for all of the intermediate vectors,
together with a constraint that the entire linear combination of all the leaf
vectors is less than or equal to the initial usage vector.
In symbols, we may let $\gr\Psi$ be the collection of leaf vectors indexed by
items in $\Delta$, and state
the constraint as $\grP \leq \sum_{\plr{x : \gr rA} \in \Delta} \gr r\gr\Psi_x$.
Seeing $\gr\Psi$ instead as a $\size\Delta \times \size\Gamma$ matrix, this
constraint is $\grP \leq \grQ\gr\Psi$, using vector-matrix multiplication.
The resulting picture is below, showing $\grP$ being split up into $\gr\Psi$,
and then each $\V$-value being constructed in a separate $\gr\Psi_i\gamma$.

@@

\begin{definition}[Usage-annotated environment]\label{def:lr-env}
  A \emph{$\V$-environment} between annotated contexts $\Gamma$ and $\Delta$
  (written $\grP\gamma$ and $\grQ\delta$, respectively, when convenient)
  is a linear map $\gr\Psi : \Ann^{\size\Delta} \to \Ann^{\size\Gamma}$ (written
  postfix) such that $\grP \leq \grQ\gr\Psi$ and for each $A$, $\grPprime$, and
  $\grQprime$ such that $\grPprime \leq \grQprime\gr\Psi$, a function from
  $\grQprime\delta \sqni A$ to $\V\,\grPprime\gamma\,A$.

  When there are multiple environments in question and $\rho$ is such an
  environment, I use the notation $\rho.\gr\Psi$ to refer to $\gr\Psi$.
  For example, $\grP \leq \grQ\plr{\rho.\gr\Psi}$.
  For the action on variables, I write $\rho(x)$, where
  $x : \grQprime\delta \sqni A$.
  The expression ``$\rho(x)$'' alone is ambiguous because of the slack in the
  usage context $\grPprime$ of the resulting value.
  Therefore, I will always make sure $\grPprime$ and $\grQprime$ clear when
  using this notation.
\end{definition}

>>>>

\begin{example}\label{ex:rec-sub}
  Take $\Ann = \plr{\mathbb N, =, 0, +, 1, \times}$, with the equality order
  chosen to avoid any concerns around subsumption of annotations.
  Then, there is an intuitionistic recursive environment (substitution) as
  follows, where $y\,z$ is the application of $y$ to $z$.
  \[
    \alr{\alr{z},\alr{y\,z}} :
    \plr{x : A, y : B \to C, z : B} \env\vdash_R \plr{B, C}
  \]
  There is also a usage-aware recursive environment
  \[
    \alr{\alr{z},\alr{y\,z}} :
    \plr{\gr0x : A, \gr2y : B \multimap C, \gr3z : B} \env\vdash_R
    \plr{\gr1B, \gr2C}.
  \]
  The latter relies on the observations that
  $\begin{pmatrix} \gr0 & \gr2 & \gr3 \end{pmatrix} =
  \begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix}
  + \begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix}$ and, on the right, that
  $\begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix} =
  \gr2\begin{pmatrix} \gr0 & \gr1 & \gr1 \end{pmatrix}$.
  Then, we have $\gr0x : A, \gr0y : B \multimap C, \gr1z : B \vdash z : B$ and
  $\gr0x : A, \gr1y : B \multimap C, \gr1z : B \vdash y\,z : C$.
\end{example}

\begin{example}\label{ex:rec-ren}
  Take $\Ann = \plr{\mathbb N, =, 0, +, 1, \times}$.
  Then, there is an intuitionistic recursive environment (renaming) as follows,
  where $\alr{-,-,-}$ abbreviates $\alr{-,\alr{-,-}}$ (matching a similar
  abbreviation in the notation of contexts).
  \[
    \alr{\alr{c},\alr{a},\alr{a}} :
    \plr{a : A, b : B, c : C, d : D} \env\sqni_R \plr{C, A, A}.
  \]
  There is also a usage-aware recursive environment
  \[
    \alr{\alr{c},\alr{a},\alr{a}} :
    \plr{\gr6a : A, \gr0b : B, \gr1c : C, \gr0d : D}
    \env\sqni_R \plr{\gr1C, \gr2A, \gr4A}.
  \]
  Intuitively, this choice of usage annotations works because the $\gr6$ $A$s on
  the left can be divided into the $\gr2 + \gr4$ $A$s on the right.
  Similarly, the $\gr0$ $B$s and $\gr0$ $D$s on the left can be discarded to
  yield none on the right.
  Note that these divisions are directed, so we cannot merge variables or
  introduce new variables on the right.
\end{example}

From \cref{ex:rec-sub}, we can see that the important usage vectors are the
initial one $\begin{pmatrix} \gr0 & \gr2 & \gr3 \end{pmatrix}$ and the usage
vectors at which terms are derived:
$\begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix}$
and $\begin{pmatrix} \gr0 & \gr1 & \gr1 \end{pmatrix}$.
I will call the latter the \emph{leaf vectors}.
The intermediate vector $\begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix}$ can
be worked out from the leaf vector
$\begin{pmatrix} \gr0 & \gr1 & \gr1 \end{pmatrix}$ and the scaling factor
$\gr2$ found in the codomain context $\gr1B, \gr2C$.
Even when the ordering on annotations is given by a non-equivalence relation
$\leq$, there is a canonical least choice for all of the intermediate vectors,
together with a constraint that the entire linear combination of all the leaf
vectors is less than or equal to the initial usage vector.
In symbols, we may let $\gr\Psi$ be the collection of leaf vectors indexed by
items in $\Delta$, and state
the constraint as $\grP \leq \sum_{\plr{x : \gr rA} \in \Delta} \gr r\gr\Psi_x$.
Seeing $\gr\Psi$ instead as a $\size\Delta \times \size\Gamma$ matrix, this
constraint is $\grP \leq \grQ\gr\Psi$, using vector-matrix multiplication.
The resulting picture is below, showing $\grP$ being split up into $\gr\Psi$,
and then each $\V$-value being constructed in a separate $\gr\Psi_i\gamma$.

@@

\begin{definition}[Usage-annotated environment]\label{def:lr-env}
  A \emph{$\V$-environment} between annotated contexts $\Gamma$ and $\Delta$
  (written $\grP\gamma$ and $\grQ\delta$, respectively, when convenient)
  is a linear map $\gr\Psi : \Ann^{\size\Delta} \to \Ann^{\size\Gamma}$ (written
  postfix) such that $\grP \leq \grQ\gr\Psi$ and for each $A$, $\grPprime$, and
  $\grQprime$ such that $\grPprime \leq \grQprime\gr\Psi$, a function from
  $\grQprime\delta \sqni A$ to $\V\,\grPprime\gamma\,A$.
\end{definition}
\begin{notation}
  When there are multiple environments in question and $\rho$ is such an
  environment, I use the notation $\rho.\gr\Psi$ to refer to $\gr\Psi$.
  For example, $\grP \leq \grQ\plr{\rho.\gr\Psi}$.
  For the action on variables, I write $\rho(x)$, where
  $x : \grQprime\delta \sqni A$.
  The expression ``$\rho(x)$'' alone is ambiguous because of the slack in the
  usage context $\grPprime$ of the resulting value.
  Therefore, I will always make sure $\grPprime$ and $\grQprime$ clear when
  using this notation.
\end{notation}

### 2023-05-02 tex/mech-related.tex
[ writing ] mostly C3 related work

\subsection{Fiore et al.}

\todo{This section is currently a random selection of paragraphs.}
Marcelo Fiore and various collaborators have a long line of work aiming for a
categorical account of variable-binding~\citep{FPT99,Fiore08,FH13,FH10,FM10}.
A recent, particularly relevant paper from this line of work is \citet{FS22},
which mechanises some of this work to obtain a framework similar in scope to
the work of \citet{AACMM21}.

Perhaps the main methodological difference between the works by
\citet{AACMM21} and \citet{FS22} is that \citet{AACMM21} start from typed
syntaxes and then give them semantics, whereas \citet{FS22} start by
characterising the category of models of a theory, and then later showing that
syntaxes form the initial model.
The latter approach means that \citeauthor{FS22} can do a significant amount of
theory not mentioning syntax in a compositional and structured style.
In contrast, the former approach is more lightweight when one wants to write
programs without a categorical specification.

I chose to base the work of this thesis on that of \citet{AACMM21} for largely
circumstantial reasons.
I only became aware of the work of \citet{FS22} when it was published, by which
time I had already completed the bulk of the work presented in this thesis.
Also, one of the authors of the former paper is my PhD supervisor, and the other
authors too were working nearby, so it was easy to discuss their work quickly
and informally.

In terms of the underlying theory, both works extend multi-sorted universal
algebra\todo{cite} with variable-binding.
However, the two extensions are subtly different.
Universal algebra already has a notion of variable, which supports renaming and
substitution, and which allows a given term to be evaluated when the free
variables of that term are assigned semantic values.
\citeauthor{AACMM21} reuse this notion of variable, and allow binding of such
variables in terms.
On the other hand, \citeauthor{FS22} recast the existing variables as
\emph{metavariables}, and introduce a new notion of (bindable) variables
separately into the syntax.
The resulting metavariables can then stand for arbitrary \emph{open} terms, and
thus each one remembers its context and requires a substitution whenever it is
used.

@@

\Citeauthor{FS22} use metavariables to form descriptions of relations over
terms, like an equational theory over the simply typed $\lambda$-calculus.
Terms have, as well as their context of variables, a context of metavariables,
and terms can contain a metavariable wherever they could contain a subterm.
There is then an operation of metavariable substitution, which substitutes terms
in the place of metavariables.
Metavariable substitution is used to instantiate the rules of described
relations/theories.
In contrast, \citeauthor{AACMM21} make do with the variables of the metalanguage
(i.e.\ Agda variables).
\todo{Congruence once and for all.}

@@

>>>>

\subsection{Second order abstract syntax}\label{sec:fiore}

Marcelo Fiore and various collaborators have a long line of work aiming for a
categorical account of variable-binding~\citep{FPT99,Fiore08,FH13,FH10,FM10}.
A recent, particularly relevant paper from this line of work is that of
\citet{FS22}, which mechanises some of this work to obtain a framework similar
in scope to the work of \citet{AACMM21}.
However, there are several differences between the resulting mechanised
frameworks, of methodological, mathematical, and technical nature.

Though \citet{AACMM21} did not state their results in categorical terms, it is
still useful to infer what the category-theoretic statement would have been, and
compare it to the statement actually given by \citet{FPT99} and \citet{FS22}.
When we do this, we see that while \citeauthor{FPT99} deal with the category of
contexts under renaming, and presheaves on that category.
This means that every model must be shown to respect renaming before touching
the framework of \citeauthor{FS22}.
Meanwhile, \citeauthor{AACMM21} make use of the interplay between the discrete
category of contexts and the category of contexts under renaming.
For example, the models (\AgdaRecord{Semantics}) of \citet{AACMM21} require only
that the family of semantic values $\V$ be shown to respect renaming, while the
fact that the result family $\C$ respects renaming follows as a corollary of the
traversal function \AgdaFunction{trav}.
\Citeauthor{FS22} make no distinction between $\V$ and $\C$, essentially only
having $\C$, but requiring it to respect renaming before getting the morphism
to that model from the initial model (the syntax).
In particular, this interplay allows us to derive renaming for terms in the way
we saw in \cref{sec:gen-syn} --- by making $\V$ the family of variables, which
trivially respects renaming.
\todo{Every model is an environment model?}

Perhaps the relative complexity of the categorical account of the work of
\citet{AACMM21} is why the authors decided not to state it in such terms.
However, it is also likely that \citeauthor{AACMM21} developed their work in
quite a different style to how \citeauthor{FPT99} did, despite arriving at
similar theories.
The work of \citet{AACMM21} is designed first and foremost to facilitate
programming language mechanisations, and thus pays a lot of attention to syntax
and traversals of it, making sure that the results compute well in Agda.
On the other hand, \citeauthor{FPT99} started from theoretical investigations
about the category of models of theories with binding, and only applied their
work to the mechanisation of programming languages in the much later work of
\citet{FS22}.

In terms of the underlying theory, both works extend multi-sorted universal
algebra\todo{cite} with variable-binding.
However, the two extensions are subtly different.
Universal algebra already has a notion of variable, which supports renaming and
substitution, and which allows a given term to be evaluated when the free
variables of that term are assigned semantic values.
\citeauthor{AACMM21} reuse this notion of variable, and allow binding of such
variables in terms.
On the other hand, \citeauthor{FS22} recast the existing variables as
\emph{metavariables}, and introduce a new notion of (bindable) variables
separately into the syntax.
The resulting metavariables can then stand for arbitrary \emph{open} terms, and
thus each one remembers its context and requires an explicit substitution
whenever it is used.

@@

\Citeauthor{FS22} use metavariables to form descriptions of relations over
terms, like an equational theory over the simply typed $\lambda$-calculus.
Terms have, as well as their context of variables, a context of metavariables,
and terms can contain a metavariable wherever they could contain a subterm.
There is then an operation of metavariable substitution, which substitutes terms
in the place of metavariables.
Metavariable substitution is used to instantiate the rules of described
relations/theories.
In contrast, \citeauthor{AACMM21} make do with the variables of the metalanguage
(i.e.\ Agda variables).

More concrete work by \citet{MMS18}, building on the approach to semantics of
\citet{ACMM17} and \cref{sec:gen-sem}, also deals with syntactic contexts into
which terms can be substituted (as part of developing notions of contextual
equivalence).
However, this work makes use of several different notions of syntactic context,
rather than just the one given naturally by metavariables in the framework of
\citeauthor{FS22}.
This suggests that more research is needed about the various roles of
metavariables before any particular approach is standardised.

@@

Despite all of these differences,
I chose to base the work of this thesis on that of \citet{AACMM21} for largely
circumstantial reasons.
I only became aware of the work of \citet{FS22} when it was published, by which
time I had already completed the bulk of the work presented in this thesis.
Also, one of the authors of the former paper is my PhD supervisor, and the other
authors too were working nearby, so it was easy to discuss their work quickly
and informally.
I think it is clear what it would mean to adapt the later work of this thesis to
a framework in the style of \citeauthor{FS22} rather than \citeauthor{AACMM21},
but it may not be obvious how usage restrictions (for ordinary variables) and
metavariables should interact.

Notably for this thesis, the work of \citet{FPT99} has been extended to
substructural syntaxes by \citet{TP06}.
This work gives a mechanism for turning a description of contexts and their
structural rules (expressed as a pseudo-monad on the 2-category of categories)
into a framework for defining substructural syntaxes, and more generally a
category of algebras of which the syntax is the initial object.
As examples, they give the untyped $\lambda$-calculus, an untyped multiplicative
linear logic, and a bunched logic.
These examples show a broad range of substructural disciplines they support ---
comparable to the work of \citet{LicataSR17} (which I discuss in
\cref{sec:semirings-conc}), and more than I discuss in this thesis.
However, they also show two of the limitations of their approach.
Firstly, this work provides no way to track types, though it should be possible
to incorporate types at the expense of complicating the categorical
constructions they use.
Secondly, it appears to be impossible to encode the syntactic forms used for the
additive connectives (i.e., the Cartesian product and coproduct) of linear
logic.
Essentially, subterms can only be combined into a larger term in the same ways
as how contexts can be appended together.
For example, in a bunched logic, contexts can be combined through both sharing
conjunction (the Cartesian product) and separating conjunction
(a monoidal product).
Correspondingly, the syntax descriptions of \citeauthor{TP06} allow for the
syntax of sharing pairs and separating pairs.
However, in the case of linear logic, contexts can only be combined via a
monoidal product, so we only get separating pairs (tensor-products) and not
sharing pairs (with-products).
\todo{Maybe this should go in the linearity chapter.}

\subsection{Substitution-based semantics}

A feature shared by the frameworks of \citet{AACMM21} and \citet{FPT99} (and
the rest of the work described in \cref{sec:fiore}) is that they are both often
concerned about how renamings act upon semantics/models.
The basic currency of both systems is presheaves on the category of contexts
under renaming, and when the user wants to produce a semantics, they must show
that the desired semantics forms such a presheaf --- amounting to showing that
the relevant type family respects renamings (in the direction that may introduce
new, unused variables to the context).

In contrast, the work of \citet{HHLM22} is based around the category of contexts
under \emph{substitution} or, more precisely, the morphisms given by semantic
environments of whatever semantics is being defined.

\begin{itemize}
  \item Hard to bootstrap.
  \item All terms have the same, countably infinite context.
\end{itemize}

\subsection{Nominal techniques}

There have been essentially two approaches when it comes to incorporating
nominal techniques into proof assistants.
The first is to develop a new foundational theory and develop a new proof
assistant on top of it.
The second is to take an existing proof assistant and provide a library,
possibly based on unsafe features.

The best-developed nominal foundational theory at the time of writing is
\emph{Fraenkel-Mostowski (FM) set theory}, as introduced by
\citet{Gabbay-thesis} and \citet{GP02}.
\citeauthor{Gabbay-thesis}'s thesis presents FM set theory, and then uses it as
the basis for a proof assistant Isabelle/FM and programming language FreshML\@.
FM is a variant of ZF set theory containing an infinite set $\mathbb A$ of
\emph{atoms} or \emph{names}, and the \emph{equivariance axiom}, stating that
every FM proposition is invariant under consistent permutation of $\mathbb A$.
In this setting, one can define the quantifier
$\reflectbox{$\mathsf N$}a.~\phi$,
read as introducing a \emph{new} or \emph{fresh} name $a$ to be used in the
proposition $\phi$.
As a material set theory, FM is unusual in that it refutes the axiom of choice.
In terms of applications to proof assistants, this means that FM is incompatible
with much existing Isabelle-based work, including anything using the Hilbert
$\varepsilon$-operator, which is used liberally in Isabelle/HOL\@.
For example, if the syntax of a programming language is formalised in FM, then
there is no formal connection to the kind of foundations in which interesting
denotational semantics have likely been formalised.

Related to nominal set theories is the recent work of \citet{Pitts23} about
theories of \emph{locally nameless sets}.
This behaves quite similarly to nominal theories, in that free variables have
names, and renaming is given by primitive operations.
In fact, \citet{Pitts23} shows that every locally nameless set is a nominal set.

The other approach --- implementing a nominal library within an existing proof
assistant --- is exemplified by the Nominal Isabelle library of \citet{Urban08}.
This work is based around having a countably infinite discrete type
\texttt{name} (which could be a type of natural numbers or strings, for example)
which is treated abstractly, and then defining liftings of permutations on
\texttt{name} to permutations on any other types involving \texttt{name}s.
Because these liftings are defined explicitly within Isabelle/HOL, the Nominal
Isabelle setup amounts to an explicit model of a nominal logic within
Isabelle/HOL.
This setup resembles what \citet{AACMM21} did within Agda, and similarly what I
do later in this thesis --- creating a library within an existing proof
assistant.
The main downside, compared to working internally to a nominal set theory, is
that the external constructions of Nominal Isabelle are more complicated, and
involve keeping track of more properties that do not come as part of the theory
the proof assistant understands.

There has also been an effort to replicate the Nominal Isabelle work in Coq by
\citet{ABW06}.
However, this work appears to have been abandoned before publication, and the
approach is unclear from the work-in-progress paper.
Nominal libraries have also been made in general-purpose programming languages,
like in Haskell with the nom package~\citep{Gabbay20}.

From a broader perspective, I claim that nominal techniques solve a subtly
different problem to that solved by the likes of Autosubst, the work of Fiore,
and the techniques discussed in detail in this chapter.
The names found in nominal techniques are more general than (names of)
variables in formal systems.
This means that nominal techniques have been applicable to problems outside
standard programming language theory\todo{Find applications}.
However, in certain ways, this is arguably not a good fit for type systems.
In a basic nominal formalisation of a syntax with binding, there is no real
distinction between the context of a term and the collection of that term's free
variables.
This, for example, leaves no natural place to put the types of variables, except
as either a partial function from the set of names to types, or from the
computed set of free variables to types.
The discrepancy becomes even more apparent in substructural systems, as we see
later, where we want tight control over the context, as opposed to the scope.

\subsection{Others}

\begin{itemize}
  \item JFDI --- \citet{BHKM12}
\end{itemize}

### 2023-05-02 tex/rensub.tex
[ writing ] mostly C3 related work

\begin{itemize}
  \item Everything presented in this chapter is perfect.
  \item But equivalence of substitutions is hard to define/work with.
  \item The process of this development gives a template for defining
    simultaneous substitution for other substructural calculi.
\end{itemize}

I have made few unforced design decisions in this chapter, so there are few
aspects I can see future work improving upon.
However, future work may want to extend the work of this chapter, in which case
there are some unanswered questions.
Principle among these, in my mind, is dealing with equivalence/equality of
environments.
Intuitionistic environments $\rho, \sigma : \Gamma \env\V \Delta$ are equal
if and only if, for each type $A$ and each variable $x : \Delta \ni A$, we have
$\rho\,x = \sigma\,x$.
This follows from what we expect of equality of functions (function
extensionality).
Usage-aware environments, on the other hand, are $\Sigma$-types --- a way of
dividing up the usage annotations of $\Gamma$, and then a function producing
$\V$-values whose usage annotations come from that division.
Equality of $\Sigma$-types is tricky --- we need to equate the first components,
then rewrite the types of the second components by this equation before equating
them.
In practice, the equations rewriting other other equations build up so much that
I have given up on a first effort to give a treatment of the equational theory
of substitutions.
Note that recursively defined environments (\cref{def:lr-rec-env}) are also
$\Sigma$-types in cases where $\Delta$ is a concatenation of contexts, so that
definition does not clearly help.

>>>>

I have made few unforced design decisions in this chapter, so there are few
aspects I can see future work improving upon.\todo{Which?}
However, future work may want to extend the work of this chapter, in which case
there are some unanswered questions.
Principal among these, in my mind, is dealing with equivalence/equality of
environments.
We want to talk about equality of environments for two related purposes.
The most immediate is that we want to develop the equational theory of renaming
and substitution --- for example, when we use \cref{thm:env-comp} to compose
substitutions, we expect that composition to be associative and unital (with
respect to \cref{thm:env-id}).
These equations of substitutions should yield equations on terms in which such
substitutions have been applied.
Slightly more abstractly, I would like to develop a theory of
\emph{quantitative multicategories}, in which multimorphisms have in their
domain a list of objects paired with usage annotations.
I would hope for $\name$ types and terms to give an example quantitative
multicategory, analogously to how Lambek calculus gives an example of an
ordinary multicategory and the simply typed $\lambda$-calculus gives an example
of a Cartesian multicategory.

Intuitionistic environments $\rho, \sigma : \Gamma \env\V \Delta$ are equal
if and only if, for each type $A$ and each variable $x : \Delta \ni A$, we have
$\rho\,x = \sigma\,x$.
This follows from what we expect of equality of functions (function
extensionality).
Usage-aware environments, on the other hand, are $\Sigma$-types --- a way of
dividing up the usage annotations of $\Gamma$, and then a function producing
$\V$-values whose usage annotations come from that division.
Equality of $\Sigma$-types is tricky --- we need to equate the first components,
then rewrite the types of the second components by this equation before equating
them.
In practice, the equations rewriting other other equations build up so much that
I have given up on a first effort to give a treatment of the equational theory
of substitutions.
Note that recursively defined environments (\cref{def:lr-rec-env}) are also
$\Sigma$-types in cases where $\Delta$ is a concatenation of contexts, so that
definition does not clearly help.

I hope that people working on substructural type systems in the future can take
inspiration from the process laid out in \cref{sec:lrkits} when working out the
appropriate notion of environment for their discipline.
Particularly, \cref{def:lr-rec-env} (the recursive definition) should serve as a
specification, if not the actual implementation, when coming across a new
substructural discipline.
As for the progression to \cref{def:lr-env}, this appears to arise from the fact
that the quantitative usage information is a refinement of the intuitionistic de
Bruijn index-based syntax.
%Being just a refinement means that the action of an environment on a variable
%is largely as it was in the intuitionistic case,

### 2023-05-02 tex/simple.tex
[ writing ] lots of background, before feedback on later chapters
[ edit ] new draft of chapters 5 and 6
[ edit ] another draft of chapters 5 and 6
[ writing ] mostly C3 related work

In this chapter, I review and justify the family of approaches usually used to
represent simple type systems inside dependently typed proof assistants.
These approaches were first presented by \citet{AR99}, who showed a way of
representing well \emph{scoped} terms in a language with polymorphic recursion,
and extended the representation to well \emph{typed} terms in a language with
dependent types.
The representation of terms relies on indexing on both a context --- giving the
types of all the free variables --- and a type for the term itself.
A basic operation on terms is \emph{simultaneous substitution}, which replaces
each variable in the context by a term in another context.
This simultaneous substitution operation fits the form of composition in a
multicategory.
Specifically, in the case of a simple intuitionistic type system, this is the
composition of a Cartesian multicategory, which allows all of the terms being
substituted in to share a context.

\section{Term representation}
\input{terms.tex}
\section{Renaming and substitution}\label{sec:kits}
\input{kits.tex}
\section{Generic semantics}
\input{gen-sem.tex}
\section{Generic syntax}
\section{Natural deduction vs sequent calculus}
\input{ndsc.tex}

>>>>

In this chapter, I review and justify the family of approaches usually used to
represent simple type systems inside dependently typed proof assistants.
These approaches were first presented by \citet{AR99}, who showed a way of
representing well \emph{scoped} terms in a language with polymorphic recursion,
and extended the representation to well \emph{typed} terms in a language with
dependent types.
The representation of terms relies on indexing on both a context --- giving the
types of all the free variables --- and a type for the term itself.
A basic operation on terms is \emph{simultaneous substitution}, which replaces
each variable in the context by a term in another context.
This simultaneous substitution operation fits the form of composition in a
multicategory.
Specifically, in the case of a simple intuitionistic type system, this is the
composition of a Cartesian multicategory, which allows all of the terms being
substituted in to share a context.
\todo{Maybe delete multicategory remark.}

This chapter almost entirely reviews prior work.
In \cref{sec:terms}, I use a presentation of the dependently typed encoding of
the simply typed $\lambda$-calculus from \citet{AR99} to set notational
conventions.
\Cref{sec:kits} presents the unpublished work of \citet{McBride05},
successively deriving simultaneous renaming and simultaneous substitution for
the terms defined in \cref{sec:terms}.
The rest of the chapter generalises the shared core of renaming and substitution
in two dimensions: in \cref{sec:gen-sem} following \citet{ACMM17} to cover
semantic traversals, and in \cref{sec:gen-syn} following \citet{AACMM21} to
cover a whole range of simply typed syntaxes with binding, rather than just
a specific syntax.
Finally, I review some related work in \cref{sec:mech-related}.

\section{Term representation}\label{sec:terms}
\input{terms.tex}
\section{Renaming and substitution}\label{sec:kits}
\input{kits.tex}
\section{Generic semantics}\label{sec:gen-sem}
\input{gen-sem.tex}
\section{Generic syntax}\label{sec:gen-syn}
\input{gen-syn.tex}
%\section{Natural deduction vs sequent calculus}
%\input{ndsc.tex}
\section{Related work}\label{sec:mech-related}
\input{mech-related.tex}
